---
title: "Basic R programming in genetics"
author: "Todsapol Techo: 777777777-7"
date: "2025-09-25"
output: html_document
---

Question 1: Create a vector containing the weights (in grams) of 10 different laboratory mice specimens.
Answer 1:
```{r}

```

Question 2: Calculate the mean, median, standard deviation, and range of these weights.
Answer 2:
```{r}

```

Question 3: Create a histogram of the mouse weights and describe the distribution.
Answer 3:
```{r}

```

Question 4: Create a data frame with columns for Sample ID, Species (choose 3 different species), Treatment (Control/Experimental), and two measurement variables relevant to your field of study.
Answer 4:
```{r}

```

Question 5: Calculate the average of your measurements grouped by Species and Treatment using appropriate R functions.
Answer 5:
```{r}

```

Question 6: Create a box plot comparing one of your measurements across treatment groups and interpret the results.
Answer 6:
```{r}

```

Question 7: Perform a t-test to compare measurements between control and experimental groups.
Answer 7:
```{r}

```

Question 8: Calculate the correlation between your two measurement variables and test for statistical significance.
Answer 8:
```{r}

```

# Use only rmarkdown format to submit